---
title: "P8106 Midterm"
author: "Brian Jo Hsuan Lee"
date: "3/26/2022"
output: pdf_document
---

Load packages

```{r, message=FALSE}
library(tidyverse)
library(corrplot)
library(caret)
library(splines)
library(mgcv)
library(earth)
```

Clean data. Consolidate the historic team names into their corresponding current names; replace NA values in weather temperature, humidity and wind speed with each of their grand averages; expand the weather detail column into boolean `dome`, `rain`, `fog`, and `snow` columns; update the spread values to those for the home team; rid seasons before 1979 due to incomplete betting line records; rid `schedule playoff` due to colinearity with the more informative `schedule week`; rid `spread_favored` and `weather details` as they are replaced by updated predictors; rid miscellaneous rows with empty values. 

```{r}
data_eda = read_csv("spreadspoke_score.csv", 
                    col_types = "iffffiiffddffiiic", 
                    col_select = c("schedule_season":"weather_detail")) %>%
  filter(
    schedule_season %in% (1979:2021)
  ) %>%
  mutate(
    schedule_season = droplevels(schedule_season),
    stadium = droplevels(stadium),
    dif = score_away - score_home, 
    weather_detail = replace(weather_detail, is.na(weather_detail), "Dry"),
    weather_detail = factor(weather_detail),
    weather_temperature = replace(weather_temperature, is.na(weather_temperature), round(mean(weather_temperature, na.rm = T), digits = 3)),
    weather_wind_mph = replace(weather_wind_mph, is.na(weather_wind_mph), round(mean(weather_wind_mph, na.rm = T), digits = 3)),
    weather_humidity = replace(weather_humidity, is.na(weather_humidity),round(mean(weather_humidity, na.rm = T), digits = 3)),
    schedule_week = fct_collapse(schedule_week, 
                                 "SuperBowl" = c("Superbowl","SuperBowl"), 
                                 "WildCard" = c("Wildcard","WildCard")),
    schedule_week = fct_relevel(schedule_week, c(1:18, "WildCard", "Division", "Conference", "SuperBowl")),
    team_home = fct_collapse(team_home, 
                             "Tennessee Titans" = c("Tennessee Titans", "Tennessee Oilers", "Houston Oilers"), 
                             "Washington Football Team" = c("Washington Football Team", "Washington Redskins"), 
                             "Las Vegas Raiders" = c("Oakland Raiders", "Los Angeles Raiders", "Las Vegas Raiders"), 
                             "Indianapolis Colts" = c("Baltimore Colts", "Indianapolis Colts"), 
                             "Los Angeles Chargers" = c("Los Angeles Chargers", "San Diego Chargers"), 
                             "Arizona Cardinals" = c("St. Louis Cardinals", "Phoenix Cardinals", "Arizona Cardinals"), 
                             "Los Angeles Rams" = c("Los Angeles Rams", "St. Louis Rams"), 
                             "New England Patriots" = c("New England Patriots", "Boston Patriots")),
    team_away = fct_collapse(team_away, 
                             "Tennessee Titans" = c("Tennessee Titans", "Tennessee Oilers", "Houston Oilers"), 
                             "Washington Football Team" = c("Washington Football Team", "Washington Redskins"), 
                             "Las Vegas Raiders" = c("Oakland Raiders", "Los Angeles Raiders", "Las Vegas Raiders"), 
                             "Indianapolis Colts" = c("Baltimore Colts", "Indianapolis Colts"), 
                             "Los Angeles Chargers" = c("Los Angeles Chargers", "San Diego Chargers"), 
                             "Arizona Cardinals" = c("St. Louis Cardinals", "Phoenix Cardinals", "Arizona Cardinals"), 
                             "Los Angeles Rams" = c("Los Angeles Rams", "St. Louis Rams"), 
                             "New England Patriots" = c("New England Patriots", "Boston Patriots")),
    team_away = fct_relevel(team_away, levels(team_home)),
    team_favorite_id = recode_factor(team_favorite_id,
                                     "MIA" = "Miami Dolphins",
                                     "TEN" = "Tennessee Titans",
                                     "LAC" = "Los Angeles Chargers",
                                     "GB" = "Green Bay Packers",
                                     "ATL" = "Atlanta Falcons",
                                     "BUF" = "Buffalo Bills",
                                     "DET" = "Detroit Lions",
                                     "PIT" = "Pittsburgh Steelers",
                                     "SF" = "San Francisco 49ers",
                                     "ARI" = "Arizona Cardinals",
                                     "WAS" = "Washington Football Team",
                                     "LAR" = "Los Angeles Rams",
                                     "CLE" = "Cleveland Browns",
                                     "DAL" = "Dallas Cowboys",
                                     "DEN" = "Denver Broncos",
                                     "MIN" = "Minnesota Vikings",
                                     "NYJ" = "New York Jets",
                                     "LVR" = "Las Vegas Raiders",
                                     "PHI" = "Philadelphia Eagles",
                                     "IND" = "Indianapolis Colts",
                                     "NE" = "New England Patriots",
                                     "KC" = "Kansas City Chiefs",
                                     "NYG" = "New York Giants",
                                     "CHI" = "Chicago Bears",
                                     "NO"= "New Orleans Saints",
                                     "CIN" = "Cincinnati Bengals",
                                     "SEA" = "Seattle Seahawks",
                                     "TB" = "Tampa Bay Buccaneers",
                                     "JAX" = "Jacksonville Jaguars",
                                     "CAR" = "Carolina Panthers",
                                     "BAL" = "Baltimore Ravens",
                                     "HOU" = "Houston Texans",
                                     .default = "None"),
    spread_home = ifelse(as.character(team_away) == as.character(team_favorite_id), abs(spread_favorite), spread_favorite),
    dome = ifelse(((as.character(weather_detail) == "DOME") | (as.character(weather_detail) == "DOME (Open Roof)")), TRUE, FALSE),
    fog = ifelse((as.character(weather_detail) == "Fog") | (as.character(weather_detail) == "Rain | Fog") | (as.character(weather_detail) == "Snow | Fog"), T, F),
    rain = ifelse((as.character(weather_detail) == "Rain") | (as.character(weather_detail) == "Rain | Fog") | (as.character(weather_detail) == "Snow | Freezing Rain"), T, F),
    snow = ifelse((as.character(weather_detail) == "Snow") | (as.character(weather_detail) == "Snow | Fog"), T, F),
  ) %>% 
  select(-score_home, -score_away, -team_favorite_id, -spread_favorite, -weather_detail) %>%
  drop_na()

data = 
  data_eda %>% 
  select(-schedule_season, -schedule_playoff)
```

Partition data into training and testing sets, and define the resampling method.

```{r}
set.seed(2022)

# partition data into training and testing sets into randomized 4:1 splits
train_index = createDataPartition(y = data$dif, p = 0.8, list = FALSE)
train_data = data[train_index, ]
train_cont_data = 
  train_data %>% 
  select(dif, over_under_line, spread_home, weather_temperature, weather_wind_mph, weather_humidity)
test_data = data[-train_index, ]

# save a expanded version of training data for exploratory data analysis
train_data_eda = data_eda[train_index, ]

# matrices of predictors 
train_pred = model.matrix(dif ~ ., train_data)[ ,-1]
train_cont_pred = model.matrix(dif ~ ., train_cont_data)[ ,-1]
test_pred = model.matrix(dif ~ ., test_data)[ ,-1]

# vectors of response
train_resp = train_data$dif
test_resp = test_data$dif

# use 2 repeats of 5-Fold CV since our dataset is rather large for our computing machines to run the typical 
# 5 repeat 10-Fold CV
ctrl = trainControl(method = "repeatedcv", repeats = 2, number = 5)
```

Visualize the distribution of our training data before analysis.

```{r, warning=FALSE}
# observe little to no correlation among continuous variables
corrplot(cor(train_cont_data), 
         method = "circle", 
         type="lower", 
         tl.cex = 0.5)

# observe no evident sign of non-linearity among continuous predictors against the response
data_theme1 = trellis.par.get()
data_theme1$plot.symbol$col = rgb(.2, .4, .2, .5)
data_theme1$plot.symbol$pch = 16
data_theme1$plot.line$col = rgb(.8, .1, .1, 1)
data_theme1$plot.line$lwd = 2
data_theme1$strip.background$col = rgb(.0, .2, .6, .2)
trellis.par.set(data_theme1)

featurePlot(train_cont_pred, train_resp, 
            plot = "scatter", labels = c("", "Score Difference (Away - Home)"),
            type = c("p"), layout = c(2, 3))

# observe relatively consistent weather information over the years
train_data_eda %>% 
  mutate(
    schedule_season = as.numeric(as.character(schedule_season))
  ) %>% 
  group_by(schedule_season) %>% 
  summarize(mean_temp = mean(weather_temperature), mean_hum = mean(weather_humidity), mean_wind = mean(weather_wind_mph)) %>% 
  pivot_longer(
    mean_temp: mean_wind,
    names_to = 'means',
    values_to = 'vals'
  ) %>%
  ggplot(aes(x = schedule_season, y = vals, color = means)) + 
  geom_line() +
  labs(
    title = "Average Humidity, Temperature and Wind Speeds across 40+ NFL Seasons",
    x = "",
    y = "Weather Values"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10),
    legend.position = "bottom"
  ) +
  scale_colour_discrete(labels = c('Humidity', 'Temperature (F)', "Wind Speed (MPH)"))

# observe similar proportions of significant games played by each team as the home team. 
team_order = 
  train_data_eda %>% 
  count(team_home, schedule_playoff) %>% 
  group_by(team_home) %>% 
  mutate(
    schedule_playoff = case_when(schedule_playoff == TRUE ~ "playoff",
                                 schedule_playoff == FALSE ~ "regular"
    )
  ) %>% 
  pivot_wider(
    names_from = schedule_playoff,
    values_from = n
  ) %>% 
  mutate(
    prop = playoff / (playoff + regular)
  ) %>% 
  arrange(desc(prop))

team_order = fct_inorder(team_order$team_home)
  
train_data_eda %>% 
  count(team_home, schedule_week) %>% 
  mutate(
    team_home = fct_relevel(team_home, levels(team_order)),
  ) %>% 
  ggplot(aes(x = team_home, y = n, fill = schedule_week)) + 
  geom_col() +
  labs(
    title = "Number of Games by Week Played by Each Team as the Home Team",
    x = "Team",
    y = "Count",
    fill = 'Schedule Week'
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.title.x = element_blank(),
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.key.size = unit(0.3, 'cm')
  )
```

Fit a k-nearest neighbor model.

```{r}
set.seed(2022)

knn_fit = train(train_pred, train_resp,
                method = "knn",
                trControl = ctrl,
                tuneGrid = expand.grid(k = seq(from = 45, to = 65, by = 1)))
```

Fit linear models

```{r, warning=FALSE}
set.seed(2022)

# Fit a multiple linear model
lm_fit = train(train_pred, train_resp,
               method = "lm",
               trControl = ctrl)

# Fit a ridge model (L2 regularization, alpha = 0)
ridge_fit = train(train_pred, train_resp,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha = 0, 
                                         lambda = exp(seq(-2, 3, length=100))),
                  preProc = c("center", "scale"),
                  trControl = ctrl)

# Fit a lasso model (L1 regularization, alpha = 1)
lasso_fit = train(train_pred, train_resp,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha = 1, 
                                         lambda = exp(seq(-3, -1, length=100))),
                  preProc = c("center", "scale"),
                  trControl = ctrl)

# Fit an elastic net model (L1 & L2 regularization, alpha = [0,1])
enet_fit = train(train_pred, train_resp,
                 method = "glmnet",
                 tuneGrid = expand.grid(alpha = seq(0, 1, length = 5),
                                        lambda = exp(seq(1, -3, length = 100))),
                 trControl = ctrl)

# Fit a principle component model
pcr_fit = train(train_pred, train_resp,
                  method = "pcr",
                  tuneGrid = data.frame(ncomp = 1:14),
                  trControl = ctrl,
                  scale = TRUE)


# Fit a partial least squares model
pls_fit = train(train_pred, train_resp,
                 method = "pls",
                 tuneGrid = data.frame(ncomp = 1:14), ## ncomp range shortened for better graphing
                 trControl = ctrl)
```

Fit nonlinear models.

```{r}
set.seed(2022)

# Fit a generalized additive model
gam_fit = train(train_pred, train_resp,
                method = "gam",
                tuneGrid = data.frame(method = "GCV.Cp", select = c(TRUE, FALSE)),
                trControl = ctrl)

# Fit a multivariate adaptive regression spline model
mars_fit = train(train_pred, train_resp,
                 method = "earth",
                 tuneGrid = expand.grid(degree = 1:3, nprune = 2:20),
                 trControl = ctrl)
```

Display the chosen model

```{r}
# ggplot(knn_fit)
# knn_fit$bestTune[1,1]
# 
# ggplot(ridge_fit, highlight = TRUE) # could xTrans = log from plot() be implemented somehow?
# # Get the optimal penalty term lambda
# ridge_fit$bestTune
# # # Count the number of non-zeroed predictors are left in the trained model
# # ridge_coef = coef(ridge_fit$finalModel, ridge_fit$bestTune$lambda)
# # ridge_coef_count = ridge_coef@p[2]; ridge_coef_count
# 
# ggplot(lasso_fit, highlight = TRUE)
# # Get the optimal penalty term lambda
# lasso_fit$bestTune
# # # Count the number of non-zeroed predictors are left in the trained model
# # lasso_coef = coef(lasso_fit$finalModel, lasso_fit$bestTune$lambda)
# # lasso_coef_count = lasso_coef@p[2]; lasso_coef_count
# 
# ggplot(enet_fit, highlight = TRUE)
# # Get tuning parameters alpha and lambda values. Alpha = 1, which matches lasso?
# enet_fit$bestTune
#
# ggplot(pcr_fit, highlight = TRUE)
# # Get the number of model components
# pcr_fit$bestTune

ggplot(pls_fit, highlight = TRUE)
# Get the number of model components
pls_fit$bestTune
```

Compile and display validation RMSE

```{r, warning=F}
resamp = resamples(list(linear = lm_fit,
                        knn = knn_fit,
                        ridge = ridge_fit,
                        lasso = lasso_fit,
                        enet = enet_fit, 
                        pcr = pcr_fit,
                        pls = pls_fit,
                        gam = gam_fit,
                        mars = mars_fit))

bwplot(
  resamples(
    list(linear = lm_fit,
         knn = knn_fit,
         ridge = ridge_fit,
         lasso = lasso_fit,
         enet = enet_fit, 
         pcr = pcr_fit,
         pls = pls_fit,
         gam = gam_fit,
         mars = mars_fit)
    ),
  metric = "RMSE")
```

Calcualte and display test RMSE

```{r, warning=F}
# calculate predicted values
knn_pred = predict(knn_fit, newdata = test_pred)
lm_pred = predict(lm_fit, newdata = test_pred)
ridge_pred = predict(ridge_fit, newdata = test_pred)
lasso_pred = predict(lasso_fit, newdata = test_pred)
enet_pred = predict(enet_fit, newdata = test_pred)
pcr_pred = predict(pcr_fit, newdata = test_pred)
pls_pred = predict(pls_fit, newdata = test_pred)
gam_pred = predict(gam_fit, newdata = test_pred)
mars_pred = predict(mars_fit, newdata = test_pred)

# join all RMSE into a dataframe
test_rmse = 
  tibble(
    knn = sqrt((test_resp - knn_pred)^2),
 linear = sqrt((test_resp - lm_pred)^2),
  ridge = sqrt((test_resp - ridge_pred)^2),
  lasso = sqrt((test_resp - lasso_pred)^2),
   enet = sqrt((test_resp - enet_pred)^2),
    pcr = sqrt((test_resp - pcr_pred)^2),
    pls = sqrt((test_resp - pls_pred)^2),
    gam = sqrt((test_resp - gam_pred)^2),
   mars = sqrt((test_resp - mars_pred)^2)
  ) %>% 
  pivot_longer(
    knn:mars,
    names_to = 'name',
    values_to = 'value'
  )

# diplay it the same way validation set RSME is displayed
ggplot(test_rmse, aes(x = name, y = value)) + 
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Test Errors Across All Fitted Models",
    y = "RMSE"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.y = element_blank()
  )
```


